---
title: "Dereplication feasibility -- all 25 participants"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

DP <- "/maps/projects/hansen_ol-AUDIT/scratch/NILU/metagenomes/assembly_annotation_wf/dereplication_pilot"

overview <- read.delim(file.path(DP, "report_overview.tsv"))
per_group <- read.delim(file.path(DP, "report_per_group.tsv"))
sweep <- read.delim(file.path(DP, "report_sweep.tsv"))
validation <- read.delim(file.path(DP, "report_validation.tsv"))
amr_redundancy <- read.delim(file.path(DP, "report_amr_redundancy.tsv"))
coassembly_unique <- read.delim(file.path(DP, "report_coassembly_unique.tsv"))
amr_coassembly_unique <- read.delim(file.path(DP, "report_amr_coassembly_unique.tsv"))
cross_participant <- read.delim(file.path(DP, "report_cross_participant.tsv"))
plasmid_amr <- read.delim(file.path(DP, "report_plasmid_amr.tsv"))
longitudinal <- read.delim(file.path(DP, "longitudinal_similarity.tsv"))

pal <- c(contigs = "#2c7fb8", orfs = "#e34a33")
```

## Context

Same-participant longitudinal metagenomics (25 participants: 19 with both a
pooled co-assembly and multiple single-sample assemblies, 6 with single-sample
assemblies only) means a lot of the sequence assembled and annotated across
samples is likely near-identical repeat, not independent new information. This
report tests that directly with `mmseqs2` clustering at two levels -- contigs
(Tier 1, feeds geNomad/MAP's contig-level tools) and ORFs/genes (Tier 2, feeds
funcscan's AMR/AMP/CAZyme tools) -- across all 25 participants, then
cross-checks the core assumption (that clustered = safe to annotate once and
propagate) against already-computed real production annotation (RGI, geNomad),
with no new annotation compute. See `dereplication_brainstorm.md` for the
original design and `dereplication_pilot_results.md` for the 2-participant
pilot this scales up.

## 1. Redundancy overview, all 25 participants

```{r overview-table}
overview %>%
  mutate(tier = factor(tier, levels = c("contigs", "orfs"))) %>%
  arrange(tier, participant) %>%
  transmute(
    Participant = participant, Tier = tier,
    `Total sequences` = format(total_seqs, big.mark = ","),
    `Redundant (count %)` = redundant_pct,
    `Redundant (bp %)` = redundant_bp_pct
  ) %>%
  kable(digits = 1)
```

```{r overview-plot, fig.height=5}
ggplot(overview, aes(x = reorder(participant, redundant_pct), y = redundant_pct, fill = tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  coord_flip() +
  labs(x = NULL, y = "Redundant sequences (% by count)", fill = "Tier",
       title = "Redundancy is high and consistent across all 25 participants") +
  theme_minimal()
```

Gene-level (Tier 2) redundancy sits at 90%+ for essentially every participant;
contig-level (Tier 1) redundancy is lower and more variable (roughly 40-45%),
consistent with contigs being longer, structurally-specific sequences while
genes are shorter, more conserved units.

## 2. Gene cluster: co-assembly vs. single-sample

```{r gene-coassembly-vs-single}
per_group %>%
  filter(tier == "orfs") %>%
  ggplot(aes(x = is_coassembly, y = redundant_pct, fill = is_coassembly)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_x_discrete(labels = c(`FALSE` = "Single-sample", `TRUE` = "Co-assembly")) +
  scale_fill_manual(values = c(`FALSE` = "#e34a33", `TRUE` = "#2c7fb8"), guide = "none") +
  labs(x = NULL, y = "Redundant genes (%)",
       title = "Co-assembly's own genes are more often redundant with single-samples\nthan the reverse (expected -- co-assembly IS the pooled single-samples)") +
  theme_minimal()
```

## 3. Contig cluster: co-assembly vs. single-sample

```{r contig-coassembly-vs-single}
per_group %>%
  filter(tier == "contigs") %>%
  ggplot(aes(x = is_coassembly, y = redundant_pct, fill = is_coassembly)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_x_discrete(labels = c(`FALSE` = "Single-sample", `TRUE` = "Co-assembly")) +
  scale_fill_manual(values = c(`FALSE` = "#e34a33", `TRUE` = "#2c7fb8"), guide = "none") +
  labs(x = NULL, y = "Redundant contigs (%)",
       title = "Same pattern at the contig level, more spread") +
  theme_minimal()
```

## 4. Co-assembly-unique contribution

Contigs/genes in a cluster with **no single-sample member at all** -- only
recoverable by pooling reads across timepoints, not by any individual
single-sample assembly.

```{r coassembly-unique-plot}
ggplot(coassembly_unique, aes(x = reorder(participant, unique_pct), y = unique_pct, fill = tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  coord_flip() +
  labs(x = NULL, y = "Co-assembly-only (%)", fill = "Tier",
       title = "What would be lost without co-assembly") +
  theme_minimal()
```

**Caveat**: a co-assembly-only cluster could reflect genuinely novel sequence
only assemblable with pooled coverage (the real win) -- or a chimeric
misassembly artifact from stitching together similar-but-distinct strain
variants across timepoints (a known risk of pooled co-assembly). This number
doesn't distinguish the two.

**Concretely, for AMR genes specifically**: `r nrow(amr_coassembly_unique)` real
resistance gene hits across all 19 co-assembly participants were found *only*
via the co-assembly -- no single-sample assembly recovered them independently.

```{r amr-coassembly-unique-table}
amr_coassembly_unique %>%
  count(aro_call, sort = TRUE) %>%
  head(15) %>%
  kable(col.names = c("ARO call", "count"),
        caption = "Most frequent co-assembly-only AMR gene calls")
```

## 5. AMR gene redundancy

```{r amr-redundancy}
total_raw <- sum(amr_redundancy$raw_hits)
total_distinct <- sum(amr_redundancy$distinct_clusters)
pct_redundant <- 100 * (total_raw - total_distinct) / total_raw
```

Across all 296 assemblies, RGI reported **`r format(total_raw, big.mark=",")`**
raw AMR gene hits, collapsing to **`r format(total_distinct, big.mark=",")`**
truly distinct AMR genes once dereplicated (**`r round(pct_redundant, 1)`%**
redundant) -- notably *lower* than the general gene-level redundancy (90%+),
meaning AMR genes show more real diversity across timepoints/samples than a
typical core gene.

```{r amr-redundancy-plot}
amr_redundancy %>%
  pivot_longer(c(raw_hits, distinct_clusters), names_to = "type", values_to = "n") %>%
  mutate(type = factor(type, levels = c("raw_hits", "distinct_clusters"),
                        labels = c("Raw RGI hits", "Distinct AMR genes"))) %>%
  ggplot(aes(x = reorder(participant, n), y = n, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = NULL,
       title = "AMR gene redundancy per participant") +
  theme_minimal()
```

## 6. Parameter sensitivity sweep

```{r sweep-table}
kable(sweep, digits = 1,
      col.names = c("Participant", "Tier", "Setting", "Redundant (count %)", "Redundant (bp %)"))
```

Tier 2 (genes) is **robust** to identity threshold (95%->99% barely moves the
number) -- the gene-level redundancy signal is real, not a loose-threshold
artifact. Tier 1 (contigs) is **highly sensitive** to `--cov-mode`: switching
from containment-aware (query-coverage) to bidirectional coverage roughly
halves the redundancy fraction -- expected, since bidirectional coverage
misses the "same region assembled to different lengths at different depths"
scenario this analysis targets. The original settings (cov-mode 2,
min-seq-id 0.95) were kept.

## 7. Propagation validation: does clustering predict identical annotation?

For clusters where >=2 members independently received a real hit from RGI
(genes) or geNomad (contigs), do they agree?

```{r validation-summary}
validation %>%
  group_by(tool) %>%
  summarise(testable = sum(testable), consistent = sum(consistent), inconsistent = sum(inconsistent)) %>%
  mutate(pct_consistent = round(100 * consistent / testable, 1)) %>%
  kable(col.names = c("Tool", "Testable clusters", "Consistent", "Inconsistent", "% consistent"))
```

```{r validation-plot}
validation %>%
  group_by(tool) %>%
  summarise(consistent = sum(consistent), inconsistent = sum(inconsistent)) %>%
  pivot_longer(c(consistent, inconsistent), names_to = "outcome", values_to = "n") %>%
  ggplot(aes(x = tool, y = n, fill = outcome)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(consistent = "#2c7fb8", inconsistent = "#e34a33")) +
  labs(x = NULL, y = "Share of testable clusters", fill = NULL,
       title = "Propagation assumption holds >96% of the time for both tools") +
  theme_minimal()
```

Both tools show strong (96-98%) agreement. The disagreements are **not
random** -- inspection of individual cases showed:

- RGI disagreements concentrate almost entirely on `vanY`/`vanT`/`vanXY` genes
  assigned to *different van-operon subtypes* (vanA vs vanB vs vanG vs vanM) --
  a known CARD/RGI database ambiguity for this highly homologous gene family,
  not a dereplication failure.
- geNomad disagreements are plasmid-vs-virus classification flips, consistent
  with genuine prophage-plasmid hybrid mobile elements the classifier finds
  genuinely borderline.

## 8. Gene cluster longitudinal similarity within participant

For every pair of a participant's own single-sample timepoints: what fraction
of gene clusters do they share (Jaccard), vs. real calendar days between them?

```{r longitudinal-plot}
ggplot(longitudinal, aes(x = days_between, y = jaccard)) +
  geom_point(aes(color = participant), alpha = 0.6, show.legend = FALSE) +
  geom_smooth(method = "loess", color = "black", se = TRUE) +
  labs(x = "Days between timepoints", y = "Jaccard similarity (shared gene clusters)",
       title = "Gene-sharing between two timepoints of the same participant decays with time") +
  theme_minimal()
```

Similarity is highest (~0.6-0.8) for timepoints ~1-2 weeks apart and drops
substantially (~0.2-0.3) for timepoints roughly a year apart -- real strain
turnover, not noise. This matters directly for the propagation design: an
"annotate once" representative may not stay valid indefinitely over a long
study and could need periodic refresh.

## 9. Within- vs. between-participant sharing (second-level clustering)

Each participant's own already-deduplicated representatives, clustered again
across all 25 participants -- a cluster with members from >1 participant means
that sequence is shared across *different people*, not just across one
person's timepoints.

```{r cross-participant-summary}
cross_participant %>%
  group_by(tier) %>%
  summarise(
    within_only = sum(n_clusters[n_participants_in_cluster == 1]),
    between = sum(n_clusters[n_participants_in_cluster > 1]),
    total = sum(n_clusters)
  ) %>%
  mutate(pct_between = round(100 * between / total, 1)) %>%
  kable(col.names = c("Tier", "Within-participant only", "Between-participant", "Total", "% between"))
```

```{r cross-participant-hist}
cross_participant %>%
  mutate(n_participants_in_cluster = pmin(n_participants_in_cluster, 10)) %>%
  group_by(tier, n_participants_in_cluster) %>%
  summarise(n_clusters = sum(n_clusters), .groups = "drop") %>%
  ggplot(aes(x = factor(n_participants_in_cluster), y = n_clusters, fill = tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  scale_y_log10() +
  labs(x = "Participants sharing a cluster (10 = 10+)", y = "Number of clusters (log scale)",
       fill = "Tier", title = "Most clusters are participant-specific, but a large tail is shared") +
  theme_minimal()
```

Genes are shared across participants far more than contigs (44% vs. 30% of
clusters span >=2 people) -- genes are smaller, more conserved units that
recombine/transfer more readily, while contigs carry more
participant/strain-specific structural context. This means the *full* cohort
dereplication opportunity is larger than the within-participant numbers above
suggest -- including real, concrete cases of the same AMR gene (e.g. `adeF`)
recurring in different, unrelated participants.

## 10. Contig cluster x plasmid classification x AMR genes

```{r plasmid-amr-table}
plasmid_amr %>%
  mutate(pct_amr = round(100 * n_amr_carrying / n_clusters, 2)) %>%
  kable(col.names = c("Category", "N clusters", "N AMR-carrying", "% AMR-carrying"))
```

```{r plasmid-amr-plot}
plasmid_amr %>%
  mutate(pct_amr = 100 * n_amr_carrying / n_clusters,
         category = factor(category, levels = c("other", "virus", "plasmid"))) %>%
  ggplot(aes(x = category, y = pct_amr, fill = category)) +
  geom_col() +
  scale_fill_manual(values = c(other = "grey70", virus = "#66c2a5", plasmid = "#fc8d62"), guide = "none") +
  labs(x = NULL, y = "AMR-carrying clusters (%)",
       title = "Plasmid-classified contig clusters carry AMR genes at ~4x background rate") +
  theme_minimal()
```

Plasmid-classified contig clusters carry an AMR gene at roughly **4x** the
background rate; virus-classified clusters show no enrichment over background
-- matches expected mobile-genetic-element biology (plasmids as a classic AMR
reservoir/vector, unlike phages), recovered here from real, cluster-deduplicated
data rather than assumed.

## Conclusions

1. **Real, large, robust redundancy at both tiers, across all 25
   participants** -- not just the original 2-participant pilot. Gene-level
   redundancy (90%+) is essentially uniform; contig-level redundancy (~40-45%)
   is more variable but still substantial.
2. **The core propagation assumption holds** (96-98% consistent for both a
   gene-level and a contig-level tool), with disagreements traceable to known,
   specific classifier ambiguities rather than random noise -- a real
   annotate-representative-and-propagate pipeline is justified.
3. **Co-assembly earns its computational cost**: recovers real AMR genes
   (607 instances) and substantial contig/gene content that no single-sample
   assembly would have, though a fraction of that may be pooling artifact
   rather than genuine novel sequence.
4. **Cross-participant sharing is real and non-trivial**, especially for
   genes (44%) -- a full cohort-wide dereplication (not just within-participant)
   would capture meaningfully more redundancy than reported here.
5. **Time matters**: gene-sharing between timepoints decays with real
   calendar time -- an eventual production pipeline should account for this
   rather than treat "annotate once" as valid indefinitely.

Not yet done: per-tool propagation-safety audit beyond RGI/geNomad (the
brainstorm's original open question), and no representative-annotation +
propagation pipeline has been built yet -- this report is the evidence base
for deciding whether to build one.
