---
title: "Dereplication feasibility -- all 25 participants"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: false
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
# echo = TRUE (was FALSE): with source hidden, "code_folding: hide"'s Code button had
# nothing to reveal -- looked broken but was working as configured. toc_float: false
# (was TRUE): renders a plain static TOC list directly in the page instead of one a
# JS library (tocify) populates at load time, since that wasn't rendering reliably in
# at least one real viewer.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(stringr)

DP <- "/maps/projects/hansen_ol-AUDIT/scratch/NILU/metagenomes/assembly_annotation_wf/dereplication_pilot"

overview <- read.delim(file.path(DP, "report_overview.tsv"))
# NOTE: is_coassembly columns were written by Python as str(bool) -- "True"/"False"
# (capitalized first letter only), which never matches R's own TRUE/FALSE when read
# via read.delim (stays a character column) -- confirmed the hard way: scale_x_discrete/
# scale_fill_manual keyed on `TRUE`/`FALSE` silently produced wrong axis labels and
# missing fill colors in this report's first two co-assembly-vs-single-sample plots
# (no error, since unmatched scale values just degrade to NA rather than fail) until a
# later chunk's `!is_coassembly` hard-errored on the same root cause and exposed it.
# Converting to real logicals immediately on load avoids this everywhere at once.
per_group <- read.delim(file.path(DP, "report_per_group.tsv")) %>%
  mutate(is_coassembly = as.logical(is_coassembly))
sweep <- read.delim(file.path(DP, "report_sweep.tsv"))
validation <- read.delim(file.path(DP, "report_validation.tsv"))
amr_redundancy <- read.delim(file.path(DP, "report_amr_redundancy.tsv"))
coassembly_unique <- read.delim(file.path(DP, "report_coassembly_unique.tsv"))
amr_coassembly_unique <- read.delim(file.path(DP, "report_amr_coassembly_unique.tsv"))
cross_participant <- read.delim(file.path(DP, "report_cross_participant.tsv"))
plasmid_amr <- read.delim(file.path(DP, "report_plasmid_amr.tsv"))
longitudinal_orfs <- read.delim(file.path(DP, "longitudinal_similarity_orfs.tsv")) %>% mutate(tier = "orfs")
longitudinal_contigs <- read.delim(file.path(DP, "longitudinal_similarity_contigs.tsv")) %>% mutate(tier = "contigs")
longitudinal <- bind_rows(longitudinal_orfs, longitudinal_contigs)
amr_within_between <- read.delim(file.path(DP, "report_amr_within_between.tsv"))
amr_plasmid_within_between <- read.delim(file.path(DP, "report_amr_plasmid_within_between.tsv"))
amr_presence <- read.delim(file.path(DP, "report_amr_presence_absence.tsv")) %>%
  mutate(is_coassembly = as.logical(is_coassembly))
amr_prevalence <- read.delim(file.path(DP, "report_amr_prevalence.tsv"))
contig_summary <- read.delim(file.path(DP, "report_contig_annotation_summary.tsv"))
two_step_validation <- read.delim(file.path(DP, "R_two_step_validation_summary.tsv"))
two_step_missed_merges <- read.delim(file.path(DP, "R_two_step_validation_missed_merges.tsv"))
mag_derep <- read.delim(file.path(DP, "R_report_mag_dereplication_summary.tsv"))
mag_instances <- read.delim(file.path(DP, "R_report_mag_instances.tsv")) %>%
  mutate(is_coassembly = as.logical(is_coassembly), date = as.Date(date))
mag_per_participant <- read.delim(file.path(DP, "R_report_mag_per_participant.tsv"))
mag_totals <- read.delim(file.path(DP, "R_report_mag_totals.tsv"))
amr_drug_class <- read.delim(file.path(DP, "R_report_amr_drug_class_lookup.tsv")) %>%
  mutate(drug_class_primary = purrr::map_chr(str_split(drug_class, ";\\s*"), 1))

pal <- c(contigs = "#2c7fb8", orfs = "#e34a33")
```

## Context

Same-participant longitudinal metagenomics (25 participants: 19 with both a
pooled co-assembly and multiple single-sample assemblies, 6 with single-sample
assemblies only) means a lot of the sequence assembled and annotated across
samples is likely near-identical repeat, not independent new information. This
report tests that directly with `mmseqs2` clustering at two levels -- contigs
(Tier 1, feeds geNomad/MAP's contig-level tools) and ORFs/genes (Tier 2, feeds
funcscan's AMR/AMP/CAZyme tools) -- across all 25 participants, then
cross-checks the core assumption (that clustered = safe to annotate once and
propagate) against already-computed real production annotation (RGI, geNomad),
with no new annotation compute. See `dereplication_brainstorm.md` for the
original design and `dereplication_pilot_results.md` for the 2-participant
pilot this scales up.

## 1. Redundancy overview, all 25 participants

```{r overview-table}
overview %>%
  mutate(tier = factor(tier, levels = c("contigs", "orfs"))) %>%
  arrange(tier, participant) %>%
  transmute(
    Participant = participant, Tier = tier,
    `Total sequences` = format(total_seqs, big.mark = ","),
    `Redundant (count %)` = redundant_pct,
    `Redundant (bp %)` = redundant_bp_pct
  ) %>%
  kable(digits = 1)
```

```{r overview-plot, fig.height=5}
ggplot(overview, aes(x = reorder(participant, redundant_pct), y = redundant_pct, fill = tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  coord_flip() +
  labs(x = NULL, y = "Redundant sequences (% by count)", fill = "Tier",
       title = "Redundancy is high and consistent across all 25 participants") +
  theme_minimal()
```

Gene-level (Tier 2) redundancy sits at 90%+ for essentially every participant;
contig-level (Tier 1) redundancy is lower and more variable (roughly 40-45%),
consistent with contigs being longer, structurally-specific sequences while
genes are shorter, more conserved units.

## 2. Gene cluster: co-assembly vs. single-sample

```{r gene-coassembly-vs-single}
per_group %>%
  filter(tier == "orfs") %>%
  ggplot(aes(x = is_coassembly, y = redundant_pct, fill = is_coassembly)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_x_discrete(labels = c(`FALSE` = "Single-sample", `TRUE` = "Co-assembly")) +
  scale_fill_manual(values = c(`FALSE` = "#e34a33", `TRUE` = "#2c7fb8"), guide = "none") +
  labs(x = NULL, y = "Redundant genes (%)",
       title = "Co-assembly's own genes are more often redundant with single-samples\nthan the reverse (expected -- co-assembly IS the pooled single-samples)") +
  theme_minimal()
```

## 3. Contig cluster: co-assembly vs. single-sample

```{r contig-coassembly-vs-single}
per_group %>%
  filter(tier == "contigs") %>%
  ggplot(aes(x = is_coassembly, y = redundant_pct, fill = is_coassembly)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_x_discrete(labels = c(`FALSE` = "Single-sample", `TRUE` = "Co-assembly")) +
  scale_fill_manual(values = c(`FALSE` = "#e34a33", `TRUE` = "#2c7fb8"), guide = "none") +
  labs(x = NULL, y = "Redundant contigs (%)",
       title = "Same pattern at the contig level, more spread") +
  theme_minimal()
```

### Both tiers together: co-assembly vs. single-sample as set overlaps

The same comparison as a set breakdown -- clusters unique to single-sample
assemblies, unique to the co-assembly, or shared between them, for both tiers.
(A true UpSet dot-matrix plot was attempted here via patchwork, but hit a
persistent axis-alignment bug combining discrete and continuous panels; with
only 2 sets being compared, a grouped bar chart shows the identical
information without the fragility.)

```{r coassembly-single-upset, fig.height=5}
sets <- read.delim(file.path(DP, "report_coassembly_vs_single_sets.tsv"))

set_levels <- c("single_only", "shared", "coassembly_only")
set_labels_full <- c(single_only = "Single-sample only", shared = "Shared", coassembly_only = "Co-assembly only")

sets_agg <- sets %>%
  group_by(tier, set) %>%
  summarise(n_clusters = sum(n_clusters), .groups = "drop") %>%
  mutate(set = factor(set, levels = set_levels))

ggplot(sets_agg, aes(x = set, y = n_clusters, fill = set)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = scales::comma(n_clusters)), vjust = -0.4, size = 3.2) +
  facet_wrap(~tier, scales = "free_y") +
  scale_x_discrete(labels = set_labels_full) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)), labels = scales::comma) +
  scale_fill_manual(values = c(single_only = "#e34a33", shared = "#66a61e", coassembly_only = "#2c7fb8"), guide = "none") +
  labs(x = NULL, y = "Clusters", title = "Co-assembly vs. single-sample: unique and shared cluster counts") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1), strip.text = element_text(face = "bold"))
```

Genes: the "shared" bar dominates (5.25M of ~7.0M pooled cluster-instances) --
consistent with the 90%+ gene-level redundancy already reported. Contigs: "single
only" dominates instead (3.04M of ~4.1M) -- most contig identity is still
sample-specific even though a meaningful fraction (both "shared" and
"co-assembly only") is not, matching the more moderate ~40-45% contig-level
redundancy already reported.

## 4. Co-assembly-unique contribution

Contigs/genes in a cluster with **no single-sample member at all** -- only
recoverable by pooling reads across timepoints, not by any individual
single-sample assembly.

```{r coassembly-unique-plot}
ggplot(coassembly_unique, aes(x = reorder(participant, unique_pct), y = unique_pct, fill = tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  coord_flip() +
  labs(x = NULL, y = "Co-assembly-only (%)", fill = "Tier",
       title = "What would be lost without co-assembly") +
  theme_minimal()
```

**Caveat**: a co-assembly-only cluster could reflect genuinely novel sequence
only assemblable with pooled coverage (the real win) -- or a chimeric
misassembly artifact from stitching together similar-but-distinct strain
variants across timepoints (a known risk of pooled co-assembly). This number
doesn't distinguish the two.

**Concretely, for AMR genes specifically**: `r nrow(amr_coassembly_unique)` real
resistance gene hits across all 19 co-assembly participants were found *only*
via the co-assembly -- no single-sample assembly recovered them independently.

```{r amr-coassembly-unique-table}
amr_coassembly_unique %>%
  count(aro_call, sort = TRUE) %>%
  head(15) %>%
  kable(col.names = c("ARO call", "count"),
        caption = "Most frequent co-assembly-only AMR gene calls")
```

## 5. AMR gene redundancy

```{r amr-redundancy}
total_raw <- sum(amr_redundancy$raw_hits)
total_distinct <- sum(amr_redundancy$distinct_clusters)
pct_redundant <- 100 * (total_raw - total_distinct) / total_raw
```

Across all 296 assemblies, RGI reported **`r format(total_raw, big.mark=",")`**
raw AMR gene hits, collapsing to **`r format(total_distinct, big.mark=",")`**
truly distinct AMR genes once dereplicated (**`r round(pct_redundant, 1)`%**
redundant) -- notably *lower* than the general gene-level redundancy (90%+),
meaning AMR genes show more real diversity across timepoints/samples than a
typical core gene.

```{r amr-redundancy-plot}
amr_redundancy %>%
  pivot_longer(c(raw_hits, distinct_clusters), names_to = "type", values_to = "n") %>%
  mutate(type = factor(type, levels = c("raw_hits", "distinct_clusters"),
                        labels = c("Raw RGI hits", "Distinct AMR genes"))) %>%
  ggplot(aes(x = reorder(participant, n), y = n, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = NULL,
       title = "AMR gene redundancy per participant") +
  theme_minimal()
```

## 6. Parameter sensitivity sweep

```{r sweep-table}
kable(sweep, digits = 1,
      col.names = c("Participant", "Tier", "Setting", "Redundant (count %)", "Redundant (bp %)"))
```

Tier 2 (genes) is **robust** to identity threshold (95%->99% barely moves the
number) -- the gene-level redundancy signal is real, not a loose-threshold
artifact. Tier 1 (contigs) is **highly sensitive** to `--cov-mode`: switching
from containment-aware (query-coverage) to bidirectional coverage roughly
halves the redundancy fraction -- expected, since bidirectional coverage
misses the "same region assembled to different lengths at different depths"
scenario this analysis targets. The original settings (cov-mode 2,
min-seq-id 0.95) were kept.

## 7. Propagation validation: does clustering predict identical annotation?

For clusters where >=2 members independently received a real hit from RGI
(genes) or geNomad (contigs), do they agree?

```{r validation-summary}
validation %>%
  group_by(tool) %>%
  summarise(testable = sum(testable), consistent = sum(consistent), inconsistent = sum(inconsistent)) %>%
  mutate(pct_consistent = round(100 * consistent / testable, 1)) %>%
  kable(col.names = c("Tool", "Testable clusters", "Consistent", "Inconsistent", "% consistent"))
```

```{r validation-plot}
validation %>%
  group_by(tool) %>%
  summarise(consistent = sum(consistent), inconsistent = sum(inconsistent)) %>%
  pivot_longer(c(consistent, inconsistent), names_to = "outcome", values_to = "n") %>%
  ggplot(aes(x = tool, y = n, fill = outcome)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(consistent = "#2c7fb8", inconsistent = "#e34a33")) +
  labs(x = NULL, y = "Share of testable clusters", fill = NULL,
       title = "Propagation assumption holds >96% of the time for both tools") +
  theme_minimal()
```

Both tools show strong (96-98%) agreement. The disagreements are **not
random** -- inspection of individual cases showed:

- RGI disagreements concentrate almost entirely on `vanY`/`vanT`/`vanXY` genes
  assigned to *different van-operon subtypes* (vanA vs vanB vs vanG vs vanM) --
  a known CARD/RGI database ambiguity for this highly homologous gene family,
  not a dereplication failure.
- geNomad disagreements are plasmid-vs-virus classification flips, consistent
  with genuine prophage-plasmid hybrid mobile elements the classifier finds
  genuinely borderline.

## 8. Gene & contig cluster longitudinal similarity within participant

For every pair of a participant's own single-sample timepoints: what fraction
of clusters do they share (Jaccard), vs. real calendar days between them --
now for both tiers.

```{r longitudinal-plot}
ggplot(longitudinal, aes(x = days_between, y = jaccard)) +
  geom_point(aes(color = participant), alpha = 0.5, show.legend = FALSE) +
  geom_smooth(method = "loess", color = "black", se = TRUE) +
  facet_wrap(~tier, scales = "free_y") +
  labs(x = "Days between timepoints", y = "Jaccard similarity (shared clusters)",
       title = "Sharing between two timepoints of the same participant decays with time,\nmuch more steeply for contigs than genes") +
  theme_minimal()
```

Gene similarity is highest (~0.6-0.8) for timepoints ~1-2 weeks apart and drops
substantially (~0.2-0.3) for timepoints roughly a year apart -- real strain
turnover, not noise. Contig similarity starts far lower (~0.14 at best) and
decays to near-zero (~0.002-0.02) -- contigs are longer, more
structurally-specific sequences that reassemble consistently far less often
than individual genes do, even within the same person over time. This matters
directly for the propagation design: an "annotate once" representative may not
stay valid indefinitely over a long study and could need periodic refresh,
more urgently for contig-level than gene-level annotation.

## 9. Within- vs. between-participant sharing (second-level clustering)

Each participant's own already-deduplicated representatives, clustered again
across all 25 participants -- a cluster with members from >1 participant means
that sequence is shared across *different people*, not just across one
person's timepoints.

```{r cross-participant-summary}
cross_participant %>%
  group_by(tier) %>%
  summarise(
    within_only = sum(n_clusters[n_participants_in_cluster == 1]),
    between = sum(n_clusters[n_participants_in_cluster > 1]),
    total = sum(n_clusters)
  ) %>%
  mutate(pct_between = round(100 * between / total, 1)) %>%
  kable(col.names = c("Tier", "Within-participant only", "Between-participant", "Total", "% between"))
```

```{r cross-participant-hist}
cross_participant %>%
  mutate(n_participants_in_cluster = pmin(n_participants_in_cluster, 10)) %>%
  group_by(tier, n_participants_in_cluster) %>%
  summarise(n_clusters = sum(n_clusters), .groups = "drop") %>%
  ggplot(aes(x = factor(n_participants_in_cluster), y = n_clusters, fill = tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  scale_y_log10() +
  labs(x = "Participants sharing a cluster (10 = 10+)", y = "Number of clusters (log scale)",
       fill = "Tier", title = "Most clusters are participant-specific, but a large tail is shared") +
  theme_minimal()
```

Genes are shared across participants far more than contigs (44% vs. 30% of
clusters span >=2 people) -- genes are smaller, more conserved units that
recombine/transfer more readily, while contigs carry more
participant/strain-specific structural context. This means the *full* cohort
dereplication opportunity is larger than the within-participant numbers above
suggest -- including real, concrete cases of the same AMR gene (e.g. `adeF`)
recurring in different, unrelated participants.

## 10. Contig cluster x plasmid classification x AMR genes

```{r plasmid-amr-table}
plasmid_amr %>%
  mutate(pct_amr = round(100 * n_amr_carrying / n_clusters, 2)) %>%
  kable(col.names = c("Category", "N clusters", "N AMR-carrying", "% AMR-carrying"))
```

```{r plasmid-amr-plot}
plasmid_amr %>%
  mutate(pct_amr = 100 * n_amr_carrying / n_clusters,
         category = factor(category, levels = c("other", "virus", "plasmid"))) %>%
  ggplot(aes(x = category, y = pct_amr, fill = category)) +
  geom_col() +
  scale_fill_manual(values = c(other = "grey70", virus = "#66c2a5", plasmid = "#fc8d62"), guide = "none") +
  labs(x = NULL, y = "AMR-carrying clusters (%)",
       title = "Plasmid-classified contig clusters carry AMR genes at ~4x background rate") +
  theme_minimal()
```

Plasmid-classified contig clusters carry an AMR gene at roughly **4x** the
background rate; virus-classified clusters show no enrichment over background
-- matches expected mobile-genetic-element biology (plasmids as a classic AMR
reservoir/vector, unlike phages), recovered here from real, cluster-deduplicated
data rather than assumed.

## 11. AMR gene dynamics: within- vs. between-participant

Section 5 counted distinct AMR genes *within* each participant separately
(summed: `r format(sum(amr_redundancy$distinct_clusters), big.mark=",")`).
Here the same AMR-carrying clusters are pushed through the second-level
(cross-participant) clustering to ask: how many are *actually* distinct once
sharing between different people is also collapsed?

```{r amr-within-between-summary}
within_total <- sum(amr_redundancy$distinct_clusters)
whole_cohort <- sum(amr_within_between$n_distinct_amr_genes)
shared_n <- sum(amr_within_between$n_distinct_amr_genes[amr_within_between$n_participants_sharing > 1])
data.frame(
  Level = c("Within-participant (summed)", "Whole-cohort (cross-participant dedup)"),
  `Distinct AMR genes` = c(within_total, whole_cohort),
  check.names = FALSE
) %>% kable()
```

`r format(within_total, big.mark=",")` within-participant-distinct AMR genes
collapse to just **`r format(whole_cohort, big.mark=",")`** once shared across
different people too -- a further
**`r round(100*(within_total-whole_cohort)/within_total, 1)`%** reduction.
**`r round(100*shared_n/whole_cohort, 1)`%** of the whole-cohort-distinct AMR
genes are found in >=2 participants.

```{r amr-within-between-hist}
ggplot(amr_within_between, aes(x = n_participants_sharing, y = n_distinct_amr_genes)) +
  geom_col(fill = "#2c7fb8") +
  labs(x = "Number of participants sharing this AMR gene", y = "Number of distinct AMR genes",
       title = "Most whole-cohort AMR genes are shared by multiple participants,\nsome present in nearly everyone") +
  theme_minimal()
```

A striking tail: some individual AMR genes recur in *most or all* 25
participants -- consistent with a shared "core" resistance gene pool
circulating in this population, distinct from the ~43% that are genuinely
participant-specific (private resistance elements, or simply not yet detected
elsewhere).

## 12. AMR plasmid dynamics: within- vs. between-participant

The same within-vs-between question, restricted to contig clusters that are
**both** geNomad-plasmid-classified **and** AMR-gene-carrying (section 10) --
is a given resistance plasmid participant-specific, or does it circulate
across people?

```{r amr-plasmid-within-between-summary}
within_p <- sum(plasmid_amr$n_amr_carrying[plasmid_amr$category == "plasmid"])
whole_cohort_p <- sum(amr_plasmid_within_between$n_distinct_clusters)
shared_p <- sum(amr_plasmid_within_between$n_distinct_clusters[amr_plasmid_within_between$n_participants_sharing > 1])
data.frame(
  Level = c("Within-participant (summed)", "Whole-cohort (cross-participant dedup)"),
  `Distinct AMR+plasmid clusters` = c(within_p, whole_cohort_p),
  check.names = FALSE
) %>% kable()
```

`r within_p` within-participant AMR-carrying plasmid clusters collapse to
**`r whole_cohort_p`** whole-cohort distinct ones. Only
**`r round(100*shared_p/whole_cohort_p, 1)`%** are shared across >=2
participants -- far less cross-participant sharing than plain AMR genes
(section 11) show. This makes sense: matching a *whole plasmid* (a longer,
multi-gene sequence) across different people's distinct bacterial populations
is a much stricter test than matching one gene.

```{r amr-plasmid-within-between-hist}
ggplot(amr_plasmid_within_between, aes(x = n_participants_sharing, y = n_distinct_clusters)) +
  geom_col(fill = "#fc8d62") +
  labs(x = "Number of participants sharing this AMR+plasmid cluster", y = "Number of distinct clusters",
       title = "Most AMR-carrying plasmids are participant-specific,\nbut a real minority circulate across multiple people") +
  theme_minimal()
```

The most-shared single AMR-carrying plasmid cluster in this cohort was found
in `r max(amr_plasmid_within_between$n_participants_sharing)` different
participants -- real evidence of a circulating resistance plasmid, not
coincidental sequence similarity.

## 13. AMR gene presence/absence: within participant, across time

Sections 5, 11-12 reported redundancy as summary percentages. Here each
whole-cohort-distinct AMR gene (from section 11) is tracked as a real
presence/absence call per sample -- "present" meaning that sample carries a
gene clustering with that AMR gene's reference hit at >=95% identity/90%
coverage, whether or not RGI happened to call a hit on that specific sample's
own copy. This is the direct view of whether a given resistance gene
persists, disappears, or reappears over a participant's own timepoints, and
how consistently it's detected across the cohort.

```{r amr-prevalence-hist}
ggplot(amr_prevalence, aes(x = n_samples)) +
  geom_histogram(binwidth = 5, fill = "#2c7fb8") +
  labs(x = "Number of samples (out of 296) an AMR gene is detected in",
       y = "Number of distinct AMR genes",
       title = "Most whole-cohort AMR genes are detected in only a handful of samples;\na long tail is near-ubiquitous") +
  theme_minimal()
```

Tiles are colored by the underlying contig's own Tier-1 (contig-level)
dereplication cluster -- "Context 1", "Context 2", etc. per gene x participant,
ranked by first appearance -- so a color change between two timepoints means
the gene reassembled on a genuinely different contig backbone, not just a
different-looking assembly of the same one. Grey tiles mark a real sample that
was collected but didn't carry that gene, so absence is never confused with
"no sample was taken" -- a genuine gap in the grid means no sample exists near
that date at all.

```{r amr-presence-heatmap, fig.height=22, fig.width=13}
top_genes <- amr_prevalence %>% slice_max(n_samples, n = 40) %>% pull(amr_gene)
gene_label <- amr_prevalence %>%
  filter(amr_gene %in% top_genes) %>%
  transmute(amr_gene, label = paste0(substr(aro_call, 1, 30), " (", amr_gene, ")"))

sample_calendar <- read.delim(file.path(DP, "report_sample_calendar.tsv")) %>%
  mutate(date = as.Date(date))

presence_top <- amr_presence %>%
  filter(amr_gene %in% top_genes, !is_coassembly, date != "")

# Rank each gene's distinct contig-cluster contexts by first appearance, per participant,
# so colors are meaningful within a row without needing raw (per-sample, non-comparable) IDs.
context_rank <- presence_top %>%
  filter(contig_cluster != "") %>%
  mutate(date = as.Date(date)) %>%
  arrange(amr_gene, participant, date) %>%
  distinct(amr_gene, participant, contig_cluster, .keep_all = TRUE) %>%
  group_by(amr_gene, participant) %>%
  mutate(context_label = paste0("Context ", pmin(row_number(), 6), ifelse(row_number() > 6, "+", ""))) %>%
  ungroup() %>%
  select(amr_gene, participant, contig_cluster, context_label)

presence_labeled <- presence_top %>%
  left_join(context_rank, by = c("amr_gene", "participant", "contig_cluster")) %>%
  transmute(amr_gene, group, participant, status = coalesce(context_label, "Context unresolved"))

relevant_participants <- unique(presence_labeled$participant)

grid <- tidyr::crossing(amr_gene = top_genes,
                         sample_calendar %>% filter(participant %in% relevant_participants)) %>%
  left_join(presence_labeled, by = c("amr_gene", "group", "participant")) %>%
  mutate(status = ifelse(is.na(status), "Sampled, gene absent", status)) %>%
  left_join(gene_label, by = "amr_gene")

status_colors <- c(
  "Sampled, gene absent" = "grey90",
  "Context 1" = "#1b9e77", "Context 2" = "#d95f02", "Context 3" = "#7570b3",
  "Context 4" = "#e7298a", "Context 5" = "#66a61e", "Context 6+" = "#e6ab02",
  "Context unresolved" = "grey50"
)

ggplot(grid, aes(x = date, y = reorder(label, match(amr_gene, top_genes)), fill = status)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = status_colors, name = "Genomic context\n(same color = same contig lineage)") +
  facet_wrap(~participant, scales = "free_x", ncol = 4) +
  labs(x = "Sample date", y = NULL,
       title = "Presence of the 40 most prevalent AMR genes, per participant, over real time",
       subtitle = "Grey = sampled, gene absent | colored = present, colored by genomic (contig) context | gap = no sample collected") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, size = 7, hjust = 1),
        axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 8, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(size = 9),
        panel.spacing = unit(0.6, "lines"))
```

Three real patterns visible directly: (1) some AMR genes, once detected in a
participant, persist across most/all of their later timepoints -- consistent
with a stable resident strain carrying that resistance gene; (2) others appear
intermittently against a background of genuine sampling (grey tiles either
side), consistent with real strain turnover/loss, or detection-limit noise
(present but at low enough abundance/coverage to sometimes fail assembly) --
this analysis alone can't distinguish the two; (3) where a persistent gene's
color changes between timepoints, the gene moved to a different contig/plasmid
backbone rather than staying on the same one -- worth flagging as a candidate
for closer inspection (real mobilization, vs. an assembly-fragmentation
artifact splitting the same true backbone differently across samples).

### Same grid, colored by contig length instead of context

Same participants, genes, dates and sample-collection background as above --
only the fill changes, from categorical genomic context to the actual length
(bp) of the underlying contig, so a change in shade between two timepoints
shows whether the gene moved to a longer/shorter assembled backbone, not just
a *different* one.

```{r amr-presence-heatmap-length, fig.height=22, fig.width=13}
grid_length <- tidyr::crossing(amr_gene = top_genes,
                                sample_calendar %>% filter(participant %in% relevant_participants)) %>%
  left_join(presence_top %>% select(amr_gene, group, participant, contig_length),
            by = c("amr_gene", "group", "participant")) %>%
  left_join(gene_label, by = "amr_gene")

ggplot(grid_length, aes(x = date, y = reorder(label, match(amr_gene, top_genes)), fill = contig_length)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(name = "Contig length (bp)", trans = "log10", na.value = "grey90",
                        labels = scales::comma) +
  facet_wrap(~participant, scales = "free_x", ncol = 4) +
  labs(x = "Sample date", y = NULL,
       title = "Same presence grid, colored by contig length",
       subtitle = "Grey = sampled, gene absent | color = length (log scale) of the contig carrying the gene | gap = no sample collected") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, size = 7, hjust = 1),
        axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 8, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(size = 9),
        panel.spacing = unit(0.6, "lines"))
```

### A closer look at genuinely mobile genes (multiple real contexts)

Most (gene, participant) pairs above only ever show one context -- the
"multi-colored row" cases where a gene really did reassemble on different
contig backbones are the interesting minority. Isolating those and faceting by
context directly (rather than color) makes each context's own timeline legible
on its own, with point size now showing contig length.

```{r amr-mobile-genes, fig.height=8}
mobile <- context_rank %>%
  group_by(amr_gene, participant) %>%
  filter(n_distinct(context_label) >= 2) %>%
  ungroup()

if (nrow(mobile) > 0) {
  mobile_presence <- presence_top %>%
    inner_join(mobile %>% select(amr_gene, participant, contig_cluster, context_label),
               by = c("amr_gene", "participant", "contig_cluster")) %>%
    left_join(gene_label, by = "amr_gene") %>%
    mutate(gene_participant = paste0(label, " -- ", participant))

  ggplot(mobile_presence, aes(x = date, y = gene_participant, size = contig_length)) +
    geom_point(aes(color = context_label), alpha = 0.8) +
    scale_color_manual(values = status_colors, guide = "none") +
    scale_size_continuous(name = "Contig length (bp)", labels = scales::comma, range = c(2, 8)) +
    facet_wrap(~context_label, ncol = 1) +
    labs(x = "Sample date", y = NULL,
         title = sprintf("%d gene x participant pairs with >=2 real genomic contexts",
                          n_distinct(paste(mobile$amr_gene, mobile$participant))),
         subtitle = "Each panel = one context rank; point size = contig length at that timepoint") +
    theme_minimal(base_size = 10) +
    theme(axis.text.y = element_text(size = 7), strip.text = element_text(face = "bold"))
} else {
  cat("No gene x participant pairs among the top 40 genes showed >=2 distinct real contexts.\n")
}
```

## 14. Contig annotation atlas: length, genes, AMR, prevalence, plasmid/viral, MAG status

Every prior contig-level section looked at one or two dimensions at a time
(redundancy, plasmid/AMR enrichment, presence/absence). Here each of the
**`r format(nrow(contig_summary), big.mark=",")`** whole-cohort-distinct
contigs that are AMR-carrying and/or geNomad plasmid/virus-classified (the
same "interesting" set section 10 uses, out of 5.48M total contig clusters --
everything else is an uninformative singleton) gets a full annotation record:
length, gene count, AMR gene count, how many samples/participants it's found
in, plasmid/virus classification, and its MAG binning record (via Binette --
"good" = completeness >=50%, contamination <10%, the standard MAG-quality
threshold; the co-assembly/single-sample Binette runs that produced this were
part of this project's separate binning workflow, not this dereplication
pipeline).

MAG binning is done **independently per sample/assembly**, so the same
whole-cohort contig can genuinely bin well in some participants' assemblies
and fail in others -- confirmed on real data (a plasmid shared across 11
participants landed in a good-quality MAG in 4 of them, and was unbinned in
the rest). A single flat "good MAG" label would hide that real heterogeneity,
so each contig's status reflects the actual fraction of its instances that
binned well, not just whether any single one did.

```{r contig-annotation-mag-summary}
contig_summary %>%
  count(mag_status) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n)) %>%
  kable(col.names = c("MAG status", "N contigs", "%"),
        caption = sprintf("MAG binning status across all %s interesting contigs", format(nrow(contig_summary), big.mark=",")))
```

Only a quarter of these AMR/plasmid/viral contigs bin *consistently* well
(>=80% of their instances); a further quarter bin well only *sometimes* --
real evidence that MAG status is often not a stable property of the sequence
itself but depends on which specific sample's assembly/coverage context it's
recovered in. Roughly half never bin well at all -- consistent with a known
binning limitation independent of the per-sample variability: plasmids and
mobile elements often have coverage/composition signals that diverge from
their host chromosome, so binners that rely on those signals systematically
under-recover them, regardless of how real or important the sequence is.

```{r contig-annotation-heatmap, fig.height=11, fig.width=11}
top_contigs <- contig_summary %>%
  arrange(desc(n_samples), desc(length)) %>%
  slice_head(n = 60) %>%
  mutate(
    contig_label = paste0(whole_cohort_id, " (", scales::comma(length), " bp)"),
    classification = case_when(
      is_plasmid & is_virus ~ "plasmid+virus",
      is_plasmid ~ "plasmid",
      is_virus ~ "virus",
      TRUE ~ "AMR only, no plasmid/virus call"
    ),
    mag_status = factor(mag_status, levels = c("consistently good MAG", "sometimes good MAG", "binned, never good quality", "never binned"))
  )
y_order <- rev(top_contigs$contig_label)
top_contigs <- top_contigs %>% mutate(contig_label = factor(contig_label, levels = y_order))

base_theme <- theme_minimal(base_size = 8) +
  theme(axis.ticks = element_blank(), panel.grid = element_blank(),
        plot.title = element_text(size = 8, hjust = 0.5),
        legend.title = element_blank(), legend.text = element_text(size = 6),
        legend.key.width = unit(0.3, "cm"))

make_continuous_col <- function(var, title, trans = "identity", show_axis = FALSE) {
  ggplot(top_contigs, aes(x = 1, y = contig_label, fill = .data[[var]])) +
    geom_tile() +
    scale_fill_viridis_c(trans = trans, labels = scales::comma) +
    labs(title = title, x = NULL, y = NULL) +
    base_theme +
    theme(axis.text.x = element_blank(),
          axis.text.y = if (show_axis) element_text(size = 5.5) else element_blank())
}
make_discrete_col <- function(var, title, palette, show_axis = FALSE) {
  ggplot(top_contigs, aes(x = 1, y = contig_label, fill = .data[[var]])) +
    geom_tile() +
    scale_fill_manual(values = palette) +
    labs(title = title, x = NULL, y = NULL) +
    base_theme +
    theme(axis.text.x = element_blank(),
          axis.text.y = if (show_axis) element_text(size = 5.5) else element_blank())
}

col_length <- make_continuous_col("length", "Length\n(bp)", trans = "log10", show_axis = TRUE)
col_genes <- make_continuous_col("n_genes", "Genes")
col_amr <- make_continuous_col("n_amr_genes", "AMR\ngenes")
col_samples <- make_continuous_col("n_samples", "Samples")
col_participants <- make_continuous_col("n_participants", "Partic-\nipants")
col_class <- make_discrete_col("classification", "Class",
  c(plasmid = "#fc8d62", virus = "#66c2a5", `plasmid+virus` = "#8da0cb", `AMR only, no plasmid/virus call` = "grey70"))
col_mag <- make_discrete_col("mag_status", "MAG",
  c(`consistently good MAG` = "#1b9e77", `sometimes good MAG` = "#66c2a5",
    `binned, never good quality` = "#e6ab02", `never binned` = "grey80"))

# NOTE: patchwork's `|`/`/` operator overloads don't dispatch correctly against this
# ggplot2 version's new S7 object system in this env ("Can't find method for generic
# `|`" -- same underlying issue hit earlier in the co-assembly/single-sample UpSet
# attempt, worked around there by dropping patchwork entirely; here wrap_plots(), the
# function-based API, works fine since it doesn't rely on operator dispatch).
patchwork::wrap_plots(list(col_length, col_genes, col_amr, col_samples, col_participants, col_class, col_mag), nrow = 1)
```

The 60 most widely-shared interesting contigs (by sample count, then length)
shown above -- real patterns worth noting: contigs found in more
samples/participants tend to be longer and more often bin consistently well
(prevalent, well-covered sequence assembles more completely and bins more
reliably); several of the most prevalent plasmid-classified contigs never
bin well regardless of prevalence, reinforcing that MAG absence here often
reflects a binning limitation rather than low confidence in the sequence
itself -- while others show the "sometimes good MAG" pattern, binning well
in some participants' assemblies and not others for the same real sequence.

### The same atlas, but as presence/absence over real time

Section 13 tracked *genes* as heatmap rows over real time; here it's the same
idea pivoted to *contigs* as the row -- one row per whole-cohort-distinct
contig (same top-60 set as above, same order), tracked across every real
sample it could possibly appear in, with the annotation side-bars from the
panel above attached directly. This is the same underlying concept the earlier
gene-level heatmap already used -- a "whole-cohort-distinct contig" is exactly
the cluster-equivalence-class dereplication itself defines, so tracking one
row across different participants/samples is consistent with, not an
extension beyond, what the clustering already represents.

```{r contig-presence-heatmap, fig.height=11, fig.width=15}
contig_presence <- read.delim(file.path(DP, "report_contig_presence_absence.tsv")) %>%
  mutate(date = as.Date(ifelse(date == "", NA, date)))

relevant_contigs <- top_contigs$whole_cohort_id
relevant_participants_c <- contig_presence %>%
  filter(whole_cohort_id %in% relevant_contigs) %>% pull(participant) %>% unique()

contig_grid <- tidyr::crossing(
    whole_cohort_id = relevant_contigs,
    sample_calendar %>% filter(participant %in% relevant_participants_c)
  ) %>%
  left_join(contig_presence %>% filter(!is_coassembly) %>%
              select(whole_cohort_id, group, local_length),
            by = c("whole_cohort_id", "group")) %>%
  mutate(present = !is.na(local_length)) %>%
  left_join(top_contigs %>% select(whole_cohort_id, contig_label), by = "whole_cohort_id")

col_presence <- ggplot(contig_grid, aes(x = date, y = factor(contig_label, levels = y_order), fill = present)) +
  geom_tile(color = "white", linewidth = 0.05) +
  scale_fill_manual(values = c(`TRUE` = "#2c7fb8", `FALSE` = "grey90"),
                     labels = c(`TRUE` = "Detected", `FALSE` = "Sampled, not detected"), name = NULL) +
  facet_wrap(~participant, scales = "free_x", nrow = 1) +
  labs(x = NULL, y = NULL, title = "Presence over real time (faceted by participant)") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, size = 5), axis.text.y = element_blank(),
        strip.text = element_text(size = 6), legend.position = "bottom")

patchwork::wrap_plots(
  list(col_length, col_genes, col_amr, col_class, col_mag, col_presence),
  widths = c(1, 1, 1, 1, 1, 10), nrow = 1
)
```

Grey tiles mean the sample exists but that specific contig wasn't detected in
it -- same "no data" vs. "genuinely absent" distinction the gene-level version
enforces. A contig row that's solid blue across most of a participant's
timepoints is a stable resident element; scattered detection against a grey
background is the contig-level equivalent of the intermittent AMR genes
already seen in section 13 -- and now directly comparable against that same
contig's length/AMR-count/classification/MAG side-bars without cross-referencing
a separate table.

### Per-participant view: AMR gene x contig, faceted by plasmid vs. MAG association

One more cut, combining sections 13 and 14 at finer grain and per participant:
each row is now a specific **(AMR gene, contig)** pair -- not just the gene
alone -- so a gene that recurs on genuinely different contig backbones gets
separate rows instead of being collapsed into one. Rows are split into two
facets by what that contig actually is: **plasmid-associated** (geNomad
plasmid-classified) vs. **MAG-associated** (landed in a good-quality bin at
least once) -- the same two categories used throughout sections 10-14, now
shown side by side the way you'd compare mobile-element-borne vs.
chromosomally-binned resistance genes directly. Presence/absence for now
(not contig count) -- the natural next step, not built yet, is separating
single-sample from co-assembly detections the same way.

```{r amr-gene-contig-participant, fig.height=9, fig.width=13}
make_gene_contig_participant_heatmap <- function(target_participant, top_n = 40) {
  p_presence <- amr_presence %>%
    filter(participant == target_participant, !is_coassembly, date != "", contig_cluster != "")

  top_pairs <- p_presence %>%
    count(amr_gene, contig_cluster, aro_call, name = "n_obs", sort = TRUE) %>%
    slice_head(n = top_n) %>%
    left_join(contig_summary %>% select(whole_cohort_id, is_plasmid, mag_status),
              by = c("contig_cluster" = "whole_cohort_id")) %>%
    mutate(
      row_label = paste0(substr(aro_call, 1, 22), " (", contig_cluster, ")"),
      facet_type = case_when(
        is_plasmid ~ "Plasmid-associated",
        mag_status %in% c("consistently good MAG", "sometimes good MAG") ~ "MAG-associated",
        TRUE ~ "Neither (AMR only / never binned)"
      )
    )

  cal <- sample_calendar %>% filter(participant == target_participant)
  presence_flag <- p_presence %>% distinct(amr_gene, contig_cluster, group) %>% mutate(present = TRUE)

  grid <- tidyr::crossing(top_pairs %>% select(amr_gene, contig_cluster, row_label, facet_type), cal) %>%
    left_join(presence_flag, by = c("amr_gene", "contig_cluster", "group")) %>%
    mutate(present = coalesce(present, FALSE))

  ggplot(grid, aes(x = date, y = row_label, fill = present)) +
    geom_tile(color = "white", linewidth = 0.1) +
    scale_fill_manual(values = c(`TRUE` = "#2c7fb8", `FALSE` = "grey90"),
                       labels = c(`TRUE` = "Detected", `FALSE` = "Sampled, not detected"), name = NULL) +
    facet_wrap(~facet_type, ncol = 1, scales = "free_y") +
    labs(x = NULL, y = NULL, title = sprintf("Participant %s", target_participant)) +
    theme_minimal(base_size = 7) +
    theme(axis.text.x = element_text(angle = 90, size = 6), axis.text.y = element_text(size = 6),
          strip.text = element_text(size = 7, face = "bold"), legend.position = "bottom")
}

# Two illustrative participants (most AMR presence records, richest real examples --
# same "show 2 example participants" pattern as the reference figure this was adapted from).
patchwork::wrap_plots(
  list(make_gene_contig_participant_heatmap("p813"), make_gene_contig_participant_heatmap("p431")),
  nrow = 1
)
```

Read each panel as: does this specific gene-on-this-specific-contig recur
across this participant's own timepoints, and is it riding on a plasmid, a
binned chromosome, or neither? A gene appearing in *both* facets (once on a
plasmid-classified contig, once on a MAG-associated one) is real evidence of
the same resistance determinant existing on two genuinely different genomic
backgrounds within the same person -- worth a closer look candidate, not
something the gene-only view in section 13 could show.

## 15. Two-step clustering validation: does the hierarchy miss real cross-participant matches?

The whole-cohort ORF/contig identity used throughout this report (e.g. section
9) is built in **two steps**: cluster within each participant first, then
cluster only the within-participant *representatives* across participants.
That second step has a known theoretical limitation -- only one representative
per within-participant cluster is ever compared cross-participant, so a real
match can be missed if that specific representative happens not to be quite
close enough, even though another member of the same cluster would have
matched fine. This bias can only ever be **conservative** (undercount sharing,
never invent it), but it had never been measured on real data.

**Direct test**: every `aro_call` (RGI's named gene/allele) that the two-step
pipeline split into >1 distinct whole-cohort ORF cluster is a candidate missed
merge. For each candidate cluster, one real representative protein sequence
was pulled and the entire pool (2,345 sequences across 112 `aro_call` values)
was re-clustered in a single step -- no hierarchy, same settings
(`--min-seq-id 0.95 -c 0.9 --cov-mode 1 --cluster-mode 2`) as the original
ORF-tier pass. If clusters that were kept separate by the two-step approach
merge back together here, that's real, direct evidence of a missed merge, not
just a theoretical possibility.

```{r two-step-validation-summary}
two_step_validation %>%
  summarise(
    aro_calls_tested = n(),
    split_confirmed = sum(!merged),
    missed_merge = sum(merged),
    total_original_clusters = sum(n_original_clusters),
    total_new_clusters = sum(n_new_clusters)
  ) %>%
  mutate(pct_missed_merge = round(100 * missed_merge / aro_calls_tested, 1),
         pct_cluster_reduction = round(100 * (total_original_clusters - total_new_clusters) / total_original_clusters, 1)) %>%
  kable(col.names = c("aro_call values tested", "Split confirmed", "Missed merge found",
                       "Total original clusters", "Total new clusters",
                       "% aro_calls affected", "% cluster reduction"))
```

```{r two-step-validation-detail}
two_step_validation %>%
  filter(merged) %>%
  arrange(desc(n_original_clusters - n_new_clusters)) %>%
  transmute(aro_call, n_original_clusters, n_new_clusters,
            clusters_collapsed = n_original_clusters - n_new_clusters) %>%
  kable(col.names = c("aro_call", "Original clusters (two-step)", "New clusters (direct)", "Clusters collapsed"))
```

```{r two-step-validation-plot}
two_step_validation %>%
  mutate(outcome = if_else(merged, "Missed merge found", "Split confirmed")) %>%
  ggplot(aes(x = n_original_clusters, y = n_new_clusters, color = outcome)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(alpha = 0.7, size = 2) +
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(values = c("Split confirmed" = "#2c7fb8", "Missed merge found" = "#e34a33")) +
  labs(x = "Original whole-cohort clusters (two-step, log scale)",
       y = "New clusters after direct single-step re-clustering (log scale)",
       color = NULL,
       title = "Most aro_call splits sit exactly on the diagonal -- real, not a hierarchy artifact") +
  theme_minimal()
```

**81 of 112 (72%) `aro_call` splits are confirmed real** -- every original
cluster stayed distinct even under direct, hierarchy-free comparison, meaning
those ORFs really are different enough to be split, exactly as the two-step
pipeline concluded. **31 of 112 (28%) show at least one real missed merge** --
but the magnitude is small: across all 112 tested calls combined, only 84 of
2,578 total original clusters (3.3%) collapse under direct re-clustering. The
effect concentrates almost entirely in the same highly diverse, hard-to-name
gene families already flagged in section 7's propagation check -- `van`-operon
genes (vanY/vanW/vanT/vanG across the vanA/B/F/G/I/M subtypes) and a handful of
single-instance cases (`tet(W)`, `qacG`, `FosA2`). These are large gene
families with many close alleles and known CARD/RGI naming ambiguity, not
representative of AMR gene clustering generally.

**Conclusion**: the two-step hierarchy's conservative bias is real but small
and concentrated, not a broad correctness problem -- it costs a few percent of
extra splitting specifically among gene families that are already ambiguous by
name, and never merges things that shouldn't be merged (the reverse error is
structurally impossible here). Scaling to the full cohort dereplication should
expect this same small, gene-family-concentrated shortfall, not a systematic
undercount across all AMR genes.

## 16. Dereplication summary: MAGs, AMR genes, plasmid contigs

Three entity types, three different real dereplication passes already run on
this cohort -- Binette bins collapsed to species-level genomes by dRep +
GTDB-Tk (section on MAG taxonomy above), AMR ORFs collapsed to whole-cohort
gene identity by this report's own two-step `mmseqs2` clustering, and geNomad
plasmid calls collapsed the same way at the contig level. This section pulls
all three together: how many distinct, non-redundant entities does this
cohort actually contain, and how are they distributed across participants
(and, where a real date exists, across time)?

### 16.1 MAGs

**Important caveat**: co-assembly and single-assembly bins were dereplicated
by dRep in two **separate, independent** runs (never compared to each other),
so the same organism recovered in both a participant's co-assembly and their
own single-sample assemblies counts as two separate "dereplicated MAGs"
below -- a real limitation of the underlying data, not double-counted by
choice.

```{r mag-summary-table}
tibble(
  metric = c("Total Binette bins (both strands, all quality)", "...good-quality (>=50% completeness, <10% contamination)",
             "Bins entering dRep", "Dereplicated MAGs (species-level genomes)",
             "...shared across >=2 participants"),
  value = c(format(mag_totals$total_bins, big.mark = ","),
            format(mag_totals$good_mags, big.mark = ","),
            format(nrow(mag_instances), big.mark = ","),
            format(nrow(mag_derep), big.mark = ","),
            sprintf("%d (%.1f%%)", sum(mag_derep$n_participants > 1), 100 * mean(mag_derep$n_participants > 1)))
) %>% kable(col.names = c("Metric", "Value"))
```

```{r mag-phylum-breakdown}
mag_derep %>%
  filter(!is.na(phylum)) %>%
  count(phylum, sort = TRUE) %>%
  mutate(phylum = forcats::fct_reorder(phylum, n)) %>%
  ggplot(aes(x = phylum, y = n)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  labs(x = NULL, y = "Dereplicated MAGs", title = "Dereplicated MAG set by GTDB phylum") +
  theme_minimal()
```

```{r mag-participant-sharing}
ggplot(mag_derep, aes(x = pmin(n_participants, 10))) +
  geom_histogram(binwidth = 1, fill = "#2c7fb8") +
  scale_x_continuous(breaks = 1:10, labels = c(1:9, "10+")) +
  labs(x = "Participants sharing a dereplicated MAG (within its own strand)",
       y = "Number of dereplicated MAGs",
       title = "Most dereplicated MAGs are participant-specific -- a real shared minority") +
  theme_minimal()
```

```{r mag-per-participant-table}
mag_per_participant %>%
  arrange(desc(n_dereplicated_mags)) %>%
  kable(col.names = c("Participant", "Dereplicated MAGs"))
```

The most cohort-shared dereplicated MAGs (by number of participants carrying
a bin from that species-level genome), tracked across real time for
single-sample instances -- same grid convention as section 13: grey means a
real sample exists but no bin from that genome was recovered in it, color
means present, colored by GTDB phylum.

```{r mag-presence-heatmap, fig.height=16, fig.width=13}
top_mags <- mag_derep %>% filter(!is.na(gtdb_taxonomy)) %>% slice_max(n_participants, n = 30, with_ties = FALSE)

mag_label <- top_mags %>%
  transmute(dereplicated_mag_id, phylum,
            label = paste0(coalesce(str_extract(gtdb_taxonomy, "(?<=s__).+"), dereplicated_mag_id)))

mag_presence <- mag_instances %>%
  filter(dereplicated_mag_id %in% top_mags$dereplicated_mag_id, !is_coassembly, !is.na(date))

relevant_participants_mag <- unique(mag_presence$participant)

grid_mag <- tidyr::crossing(dereplicated_mag_id = top_mags$dereplicated_mag_id,
                             sample_calendar %>% filter(participant %in% relevant_participants_mag)) %>%
  left_join(mag_presence %>% distinct(dereplicated_mag_id, group, participant) %>% mutate(present = TRUE),
            by = c("dereplicated_mag_id", "group", "participant")) %>%
  left_join(mag_label, by = "dereplicated_mag_id") %>%
  mutate(fill_class = ifelse(is.na(present), "Sampled, MAG absent", phylum))

phylum_levels <- sort(unique(mag_label$phylum))
phylum_colors <- setNames(scales::hue_pal()(length(phylum_levels)), phylum_levels)
mag_fill_colors <- c(phylum_colors, "Sampled, MAG absent" = "grey90")

ggplot(grid_mag, aes(x = date, y = reorder(label, match(phylum, phylum_levels)), fill = fill_class)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = mag_fill_colors, name = "Phylum (present) /\nabsent") +
  facet_wrap(~participant, scales = "free_x", ncol = 4) +
  labs(x = "Sample date", y = NULL,
       title = "Presence of the 30 most cohort-shared dereplicated MAGs, per participant, over real time",
       subtitle = "Grey = sampled, MAG absent | colored = present, colored by GTDB phylum | gap = no sample collected") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, size = 6, hjust = 1),
        axis.text.y = element_text(size = 6),
        strip.text = element_text(size = 8, face = "bold"),
        legend.position = "bottom", legend.title = element_text(size = 9),
        panel.spacing = unit(0.6, "lines"))
```

### 16.2 AMR genes, with real antibiotic class

`aro_call` names (`vanG`, `tetB(P)`, ...) don't say what drug they confer
resistance to -- RGI's own raw output does (`Drug Class`, `Resistance
Mechanism`, `AMR Gene Family` columns), joined in here for the first time.

```{r amr-drug-class-table}
amr_prevalence %>%
  left_join(amr_drug_class, by = "aro_call") %>%
  count(drug_class_primary, name = "n_distinct_amr_genes", sort = TRUE) %>%
  kable(col.names = c("Drug class", "Distinct whole-cohort AMR genes"))
```

Same top-40 whole-cohort AMR genes, same per-participant/date grid as
section 13 -- but now colored by antibiotic drug class instead of genomic
context, with rows grouped by class, directly answering "which resistances,
to which drugs, are present in whom and when."

```{r amr-drug-class-heatmap, fig.height=22, fig.width=13}
gene_label_dc <- amr_prevalence %>%
  filter(amr_gene %in% top_genes) %>%
  left_join(amr_drug_class, by = "aro_call") %>%
  transmute(amr_gene, drug_class_primary = coalesce(drug_class_primary, "unknown"),
            label = paste0(substr(aro_call, 1, 28), " (", amr_gene, ")"))

grid_dc <- tidyr::crossing(amr_gene = top_genes,
                            sample_calendar %>% filter(participant %in% relevant_participants)) %>%
  left_join(presence_top %>% distinct(amr_gene, group, participant) %>% mutate(present = TRUE),
            by = c("amr_gene", "group", "participant")) %>%
  left_join(gene_label_dc, by = "amr_gene") %>%
  mutate(fill_class = ifelse(is.na(present), "Sampled, gene absent", drug_class_primary))

dc_levels <- sort(unique(gene_label_dc$drug_class_primary))
dc_colors <- setNames(scales::hue_pal()(length(dc_levels)), dc_levels)
dc_fill_colors <- c(dc_colors, "Sampled, gene absent" = "grey90")

ggplot(grid_dc, aes(x = date, y = reorder(label, match(drug_class_primary, dc_levels)), fill = fill_class)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = dc_fill_colors, name = "Drug class (present) /\nabsent") +
  facet_wrap(~participant, scales = "free_x", ncol = 4) +
  labs(x = "Sample date", y = NULL,
       title = "Presence of the 40 most prevalent whole-cohort AMR genes, colored by antibiotic drug class",
       subtitle = "Grey = sampled, gene absent | colored = present, colored by drug class | rows grouped by class | gap = no sample collected") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, size = 7, hjust = 1),
        axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 8, face = "bold"),
        legend.position = "bottom", legend.title = element_text(size = 9),
        panel.spacing = unit(0.6, "lines"))
```

### 16.3 geNomad plasmid contigs

```{r plasmid-derep-table}
plasmid_derep <- contig_summary %>% filter(is_plasmid)

tibble(
  metric = c("Dereplicated plasmid contigs (whole-cohort)", "...shared across >=2 participants",
             "...also AMR-carrying"),
  value = c(format(nrow(plasmid_derep), big.mark = ","),
            sprintf("%d (%.1f%%)", sum(plasmid_derep$n_participants > 1), 100 * mean(plasmid_derep$n_participants > 1)),
            sprintf("%d (%.1f%%)", sum(plasmid_derep$has_amr), 100 * mean(plasmid_derep$has_amr)))
) %>% kable(col.names = c("Metric", "Value"))
```

```{r plasmid-derep-hist}
ggplot(plasmid_derep, aes(x = pmin(n_participants, 10))) +
  geom_histogram(binwidth = 1, fill = "#e34a33") +
  scale_x_continuous(breaks = 1:10, labels = c(1:9, "10+")) +
  labs(x = "Participants sharing a dereplicated plasmid contig",
       y = "Number of dereplicated plasmid contigs",
       title = "Most dereplicated plasmid contigs are participant-specific too") +
  theme_minimal()
```

## Conclusions

1. **Real, large, robust redundancy at both tiers, across all 25
   participants** -- not just the original 2-participant pilot. Gene-level
   redundancy (90%+) is essentially uniform; contig-level redundancy (~40-45%)
   is more variable but still substantial.
2. **The core propagation assumption holds** (96-98% consistent for both a
   gene-level and a contig-level tool), with disagreements traceable to known,
   specific classifier ambiguities rather than random noise -- a real
   annotate-representative-and-propagate pipeline is justified.
3. **Co-assembly earns its computational cost**: recovers real AMR genes
   (607 instances) and substantial contig/gene content that no single-sample
   assembly would have, though a fraction of that may be pooling artifact
   rather than genuine novel sequence.
4. **Cross-participant sharing is real and non-trivial**, especially for
   genes (44%) -- a full cohort-wide dereplication (not just within-participant)
   would capture meaningfully more redundancy than reported here.
5. **Time matters**: gene-sharing between timepoints decays with real
   calendar time, far more steeply for contigs than genes -- an eventual
   production pipeline should account for this rather than treat "annotate
   once" as valid indefinitely.
6. **AMR genes show a shared "core" resistance pool**: whole-cohort dedup
   collapses 12,227 within-participant-distinct AMR genes to just 2,497 --
   57% of those are found in multiple participants, some in nearly all 25.
7. **AMR-carrying plasmids are mostly participant-specific, but not always**:
   only 12% are shared across people (vs. 57% for plain AMR genes) -- matching
   a whole plasmid is a stricter test than matching one gene -- but the most
   widely shared one was found in 7 different participants, real evidence of
   a circulating resistance plasmid in this cohort.
8. **AMR gene presence/absence has real temporal structure**, visible at the
   individual-gene level: some resistance genes persist across a
   participant's later timepoints once detected (consistent with a stable
   resident strain), others appear intermittently (strain turnover, or
   detection-limit noise -- not distinguishable from clustering alone).
9. **MAG binning success is often sample-specific, not a stable property of
   the sequence**: only ~25% of AMR/plasmid/viral contigs bin consistently
   well across every instance; another ~26% bin well only in *some*
   participants' assemblies (confirmed on a real plasmid: good MAG in 4 of
   11 participants, unbinned in the rest); ~48% never bin well at all,
   consistent with a known limitation (plasmids' coverage/composition
   diverging from their host) rather than low confidence in the sequence
   itself. Any downstream analysis using only binned/MAG-level data should
   expect to systematically miss a real, non-trivial share of AMR/plasmid
   content -- and inconsistently so across participants for the same
   sequence.

10. **The two-step (within-then-across-participant) clustering hierarchy's
    known conservative bias is real but small**: a direct hierarchy-free
    re-clustering test confirmed 72% of its `aro_call` splits as genuine, and
    found the remaining 28% collapse only 3.3% of total clusters overall --
    concentrated almost entirely in already-known-ambiguous `van`-operon gene
    families (section 7). It never merges things that shouldn't be merged.

Not yet done: per-tool propagation-safety audit beyond RGI/geNomad (the
brainstorm's original open question), and no representative-annotation +
propagation pipeline has been built yet -- this report is the evidence base
for deciding whether to build one.
